Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-06-23 article.Rmd
AHR section revisions

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession, scholarly research, and focus on two types of scholarly output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the American Historical Review reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. Once published, the impact of a book is often shaped by its reviews in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the *AHR* publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." [American Historical Review n.d.] The *AHR* is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, longevity, and reputation of the *AHR* makes it an ideal proxy to study gatekeeping within the historical discipline.

@@

Scraping the table-of-contents of every issue of the American Historical Review results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could in turn use to plot the proportion of female and male authors over time. The temporal trajectory of gender representations roughly resembles that of history dissertations: women began making inroads in the late 1960s and have made steady gains over the past four decades. By the twenty-first century the proportion of female authors reviewed in the *AHR* had more than tripled.

@@

But a closer look shows important differences between the proportion of women writing history dissertations and the proportion of women appearing in the *AHR*. Although both have trended upwards since the 1970s the slope is a lot steeper for dissertations than it is for the *AHR*. Female historians have very nearly closed the gap in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of what happens from that point onwards. In book reviews published in the *AHR* male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over into the pages of the historical profession's leading journal.^[The chart below includes only authors whose gender we successfully inferred; it does not include authors of unknown gender as in the previous charts.]

@@

The path from a dissertation to a monograph to a book review in the *American Historical Association* isn't automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] But fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the successively widening gap. A 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for research. [Townsend 2012] This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. Their male colleagues, meanwhile, enjoy more time to publish and receive more accolades when they do. As of 2014, the Pulitzer Prize for History has been awarded to exactly one female historian in the past nineteen years. The Bancroft Prize in American History has been awarded to almost four times as many male historians as female historians over that same period. ["Bancroft Prize"] The *American Historical Review* is, unfortunately, far from alone in its gatekeeping practices.

>>>>

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession, scholarly research, and focus on two types of scholarly output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the *American Historical Review* reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

@@

But the dissertation is only the first major piece of scholarship produced by an academic historian. The second usually centers on the writing of an academic monograph to be read and evaluated by their peers. These monographs remain the "coin of the realm" for many historians, and the value of that coin often depends on it being reviewed in academic journals. To study how gender structures the monograph reviews, we turn to one of the leading journals in the historical profession: the *American Historical Review*, the flagship journal of the American Historical Association. The *AHR* publishes roughly one thousand total book reviews spread across five issues each year, covering (in its own words) "every major field of historical study." [American Historical Review n.d.] The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, longevity, and reputation of the *AHR* makes it an ideal proxy to study gendered gatekeeping related to historical monographs.

@@

Scraping the table-of-contents of every issue of the *American Historical Review* results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could in turn use to plot the proportion of female and male authors over time. The temporal trajectory of gender representation in the journal’s reviews roughly resembles that of history dissertations: women began making inroads in the late 1960s and continue to make steady gains over the next four decades. By the twenty-first century, the proportion of female authors reviewed in the *AHR* had more than tripled.

@@

But a closer look shows substantial differences between the proportion of women writing history dissertations and the proportion of women appearing in the *AHR*. Although both have trended upwards since the 1970s, the slope of that line is a lot steeper for dissertations than it is for the *AHR*. Female historians have very nearly closed the gap in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of authoring monographs that are then reviewed by the discipline’s leading academic journal. In book reviews published in the *AHR*, male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over into the pages of the *American Historical Review*.^[The chart below includes only authors whose gender we successfully inferred; it does not include authors of unknown gender as in the previous charts.]

@@

Not every historian wants or seeks a career path that neatly moves from writing a dissertation to authoring a monograph to being reviewed in the *American Historical Review*. But for those that do, each of these subsequent stages represents a winnowing process: not all graduate students produce dissertations, not all PhD’s write books, not all books are submitted to the *American Historical Review*, and not all submissions are accepted by the *AHR* editors for review. Nearly equal numbers of female and male historians make it through the initial stage of writing a dissertation. But in the final round of gatekeeping – a monograph review in the *AHR* – there are nearly twice as many men as women.

Without more information it is all but impossible to establish a causal reason for this pattern. Instead, a murky concoction of factors likely contributes to the different gender gaps between dissertations and monograph reviews in the AHR. For one, dissertations and monographs are frequently written not only at different points in a person’s career, but also their family life. A 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for research. [Townsend 2012] This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. Their male colleagues, meanwhile, enjoy more time to publish their research.

***left off revising here***

The path from a dissertation to a monograph to a book review in the *American Historical Association* isn't automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the *AHR* receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] But fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the successively widening gap. A 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for research. [Townsend 2012] This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. Their male colleagues, meanwhile, enjoy more time to publish and receive more accolades when they do. As of 2014, the Pulitzer Prize for History has been awarded to exactly one female historian in the past nineteen years. The Bancroft Prize in American History has been awarded to almost four times as many male historians as female historians over that same period. ["Bancroft Prize"] The *American Historical Review* is, unfortunately, far from alone in its gatekeeping practices.
